---
title: "LAB #2: Solving Differential Equations"
subtitle: MATH B210
author: Michael Pham
date: 9/25/2023
output: 
  html_document:
    theme: cosmo
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(ggplot2)
```

## Instructions
1. Read and work through the Lab 2 document.
2. Complete the exercises below.
3. Delete these instructions from your final R Markdown document.

## Exercise 1

Consider the differential equation $\frac{dy}{dt} = −ty$ with initial condition $y(−1) = −1$. Using Euler’s method, plot the solution from $t = −1 \text{ to } 3$ with the number of steps $n = 4,~8,~16,~32$, and $64$. Assign each solution a different color and/or line type, and include a legend. At which number of steps do you begin to see an accurate result? What's your criteria for accuracy?
```{r}
t <- -1
y <- -1
df <- data.frame(x = numeric(0), y = numeric(0), set = numeric(0))

diff_eq <- function(t, y) {
  result <- -t * y
  return(result)
}

euler_method <- function(steps, diff_func) {
  delta_t <- 4/steps
  t_vector <- c(t)
  y_vector <- c(y)
  for (i in 0:(steps-1)) {
    y <- y + diff_func(t, y) * delta_t
    t <- t + delta_t

    t_vector <- c(t_vector, t)
    y_vector <- c(y_vector, y)
  }

  # print(t_vector)
  print(y_vector)

  step_size_vector <- rep(steps, length(t_vector))
  df <<- rbind(df, data.frame(x = t_vector, y = y_vector, steps = step_size_vector))

}

lst <- list(4,8,16,32,64)
for (element in lst) {
  euler_method(element, diff_eq)
}

#ykplusone + yk + f(tk, yk)delta(t)
```
```{r}
ggplot(df, aes(x, y, color = factor(steps))) +
  geom_point(size = 5) +
  geom_line(aes(group = steps), size = 1) +
  labs(x = "t", y = "y") +
  scale_color_manual(values = c("4" = "red", "8" = "blue", "16" = "green", "32" = "purple", "64" = "orange")) +
  theme_minimal()
```
```{r}
#At steps = 16, the solutions look more accuracte. There's more curvature to the solution, whereas before with only 4 steps, there's more defined over and underestimation of the actual solution.
```

## Exercise 2

A population of fish grows based on the logistic model, but is constantly being fished at a rate $h$. This is called the <i>logistic model with constant harvesting</i>. The change in population is given by
\[
\frac{dp}{dt} = p\left(1-\frac{p}{N}\right) -h
\]
where $N$ is the carrying capacity and $h$ is the rate of harvesting. The variables are the population $p$ of fish and the time $t$ in years.

Suppose that $N=100$ and $h=25$. Given an initial population of $p_0 = 80$, what happens to the fish population in the next 10 years? Use the Runge-Kutta method to find a numerical solution. Provide a plot and explain what happens to the population.
```{r}
diff_eq2 <- function(t, p){
  return ( p*(1 - p/100) - 25 )
}

rk4_method <- function(t, p, steps, interval, diff_func){
  t_vector <- vector("numeric", length = 0)
  p_vector <- vector("numeric", length = 0)
  
  delta_t = interval/steps
  
  for (i in 0:(steps-1)) {
    t_vector <- c(t_vector, t)

    t_mid <- t + delta_t/2

    mk <- diff_func(t, p)
    nk <- diff_func(t_mid, p + mk*(delta_t/2))
    qk <- diff_func(t_mid, p + nk*(delta_t/2))

    t <- t + delta_t
    pk <- diff_func(t, p + qk*delta_t)

    p <- p + ((mk + 2*nk + 2*qk + pk)/6) * delta_t

    p_vector <- c(p_vector, p)
  }
  return (list(t = t_vector, p = p_vector))
}

list <- rk4_method(0, 80, 100, 10, diff_eq2)
print(list$t)
print(list$p)
```

```{r}
data <- data.frame(t = list$t, p = list$p)
  ggplot(data, aes(x = t, y = p)) +
    geom_point() +
    geom_line() +
    labs(x = "Time (years)", y = "Population", title = "Scatter Plot of Fish Population Solution")
```
```{r}
#The population decreases but at a slower and slower rate
```

## Exercise 3

To protect the fish population, fishers are restricted to harvesting the fish  in a particular season. Therefore, the rate of harvesting is periodic. The equation for the growth rate becomes
\[
\frac{dp}{dt} = p\left(1-\frac{p}{N}\right) -h\frac{1+\sin bt}{2}
\]
Suppose that $N = 100$, $h = 5$, and $b = 1$. Use the Runge-Kutta method to find a numerical solution. Plot the solution to this differential equation for $p_0 = 80$. What is the long-term behavior of the fish population? 
```{r}
diff_eq3 <- function(t, p){
  return ( p*(1 - p/100) - 5*((1+sin(t))/2) )
}

list <- rk4_method(0, 80, 100, 10, diff_eq3)
print(list$t)
print(list$p)

data <- data.frame(t = list$t, p = list$p)
  ggplot(data, aes(x = t, y = p)) +
    geom_point() +
    geom_line() +
    labs(x = "Time (years)", y = "Population", title = "Scatter Plot of Fish Population Solution")

```

```{r}
#The population increases but then the sinosoidal rate of change in the second term starts to take over after about 5 years when the population gets close to the carrying capacity. After this, the fish population will fluctuate sinosoidally in the ~96 to ~99 range.

```

## Exercise 4

Consider the initial-value problem $\displaystyle\frac{dy}{dt}=\frac{1}{(y+1)(t-2)}$ with $y(-2)=0$. The analytical solution for this IVP is
\[
y(t) = -1 + \sqrt{1+\ln \left(\frac{(t-2)^2}{16}\right)}
\]

(a) Plot the analytical solution for the IVP over the interval $t=-2$ to $t=2$. Add to the plot numerical solutions using Euler's Method, with the number of steps $n=64,~128,~256$. Adjust the y-axis limits as needed.
```{r}
diff_eq4 <- function(t,y) {
  return( 1/((y+1)*(t-2)) )
}

solution <- function(t) {
  return ( -1 + sqrt(1 + log(((t-2)^2))/16) )
}

euler_method2 <- function(t, y, steps, end, diff_func) {
  step_size <- (end-t)/steps
  
  t_vector <- vector("numeric", length = 0)
  y_vector <- vector("numeric", length = 0)

  for (i in seq(t, end, by = step_size)) {
    y_vector <- c(y_vector, y)
    t_vector <- c(t_vector, t)

    y <- y + step_size * diff_func(t, y)
    t <- t + step_size
  }
  return (list(y = y_vector, t = t_vector))

}



lst <- list(4,8,16,32,64)
df <- data.frame(x = numeric(0), y = numeric(0), set = numeric(0))

for (element in lst) {
  list <- euler_method2(-2, 0, element, 2, diff_eq4)
  df <- rbind(df, data.frame(x = list$t, y = list$y, set = rep(element, length(list$t))))
}


t <- seq(-2, 2, by = 0.01)
y <- sapply(t, solution)

df <- rbind(df, data.frame(x = t, y = y, set = rep("real", length(t))))


print(df)

ggplot(df, aes(x, y, color = factor(set))) +
  geom_point(size = 5) +
  geom_line(aes(group = set), size = 1) +
  labs(x = "t", y = "y") +
  scale_color_manual(values = c("4" = "red", "8" = "blue", "16" = "green", "32" = "purple", "64" = "orange", "real" = "violet")) +
  theme_minimal()
```
(b) Again, plot the analytical solution over the interval $t=-2$ to $t=2$. Now add, the 4th-order Runga-Kutta solutions using $n=100$ and $200$. Adjust the y-axis limits as needed.
```{r}

list <- rk4_method(-2, 0, 100, 4, diff_eq4)
df <- rbind(df, data.frame(x = list$t, y = list$p, set = rep("rk4-100", length(list$t))))
list <- rk4_method(-2, 0, 200, 4, diff_eq4)
df <- rbind(df, data.frame(x = list$t, y = list$p, set = rep("rk4-200", length(list$t))))

ggplot(df, aes(x, y, color = factor(set))) +
  geom_point(size = 5) +
  geom_line(aes(group = set), size = 1) +
  labs(x = "t", y = "y") +
  scale_y_continuous(limits = c(-5, 1)) + 
  scale_color_manual(values = c("4" = "red", "8" = "blue", "16" = "green", "32" = "purple", "64" = "orange", "real" = "violet", "rk4-100" = "#1f9ab9", "rk4-200" = "#2a6776")) +
  theme_minimal()
```
(c) Comment on your observations of the previous two parts. Do you have any explanation for your observations?

```{r}
#Even though the given solution is a pretty continous line, the solution obtained from Euler's and RK jump all over the place with big discrete jumps rather than smooth derivative changes. For Euler's a high step count of 64 resulted in the solution most closely matching the real provided solution. For RK method, a bigger step size resulted in a less accurate solution graph. Also, RK with step size of 100 did not do better than Euler with only a step size of 64.
```