---
title: "LAB #2: Solving Differential Equations"
subtitle: MATH B210
author: Michael Pham
date: 9/25/2023
output: 
  html_document:
    theme: cosmo
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(ggplot2)
```

## Instructions
1. Read and work through the Lab 2 document.
2. Complete the exercises below.
3. Delete these instructions from your final R Markdown document.

## Exercise 1

Consider the differential equation $\frac{dy}{dt} = −ty$ with initial condition $y(−1) = −1$. Using Euler’s method, plot the solution from $t = −1 \text{ to } 3$ with the number of steps $n = 4,~8,~16,~32$, and $64$. Assign each solution a different color and/or line type, and include a legend. At which number of steps do you begin to see an accurate result? What's your criteria for accuracy?
```{r}
t <- -1
y <- -1
df <- data.frame(x = numeric(0), y = numeric(0), set = numeric(0))

diff_eq <- function(t, y) {
  result <- -t * y
  return(result)
}

euler_method <- function(steps) {
  delta_t <- 4/steps
  t_vector <- c(t)
  y_vector <- c(y)
  for (i in 0:(steps-1)) {
    y <- y + diff_eq(t, y) * delta_t
    t <- t + delta_t

    t_vector <- c(t_vector, t)
    y_vector <- c(y_vector, y)
  }

  # print(t_vector)
  print(y_vector)

  step_size_vector <- rep(steps, length(t_vector))
  df <<- rbind(df, data.frame(x = t_vector, y = y_vector, steps = step_size_vector))

}

lst <- list(4,8,16,32,64)
for (element in lst) {
  euler_method(element)
}

#ykplusone + yk + f(tk, yk)delta(t)
```
```{r}
ggplot(df, aes(x, y, color = factor(steps))) +
  geom_point(size = 5) +
  geom_line(aes(group = steps), size = 1) +
  labs(x = "t", y = "y") +
  scale_color_manual(values = c("4" = "red", "8" = "blue", "16" = "green", "32" = "purple", "64" = "orange")) +
  theme_minimal()
```
```{r}
#At steps = 16, the solutions look more accuracte. There's more curvature to the solution, whereas before with only 4 steps, there's more defined over and underestimation of the actual solution.
```

## Exercise 2

A population of fish grows based on the logistic model, but is constantly being fished at a rate $h$. This is called the <i>logistic model with constant harvesting</i>. The change in population is given by
\[
\frac{dp}{dt} = p\left(1-\frac{p}{N}\right) -h
\]
where $N$ is the carrying capacity and $h$ is the rate of harvesting. The variables are the population $p$ of fish and the time $t$ in years.

Suppose that $N=100$ and $h=25$. Given an initial population of $p_0 = 80$, what happens to the fish population in the next 10 years? Use the Runge-Kutta method to find a numerical solution. Provide a plot and explain what happens to the population.
```{r}
n = 100
h = 25

diff_eq2 <- function(t, p){
  return ( p*(1 - p/n) - h )
}

rk4_method <- function(t0, p0){

}

rk4_method(0, 80)
```


## Exercise 3

To protect the fish population, fishers are restricted to harvesting the fish  in a particular season. Therefore, the rate of harvesting is periodic. The equation for the growth rate becomes
\[
\frac{dp}{dt} = p\left(1-\frac{p}{N}\right) -h\frac{1+\sin bt}{2}
\]
Suppose that $N = 100$, $h = 5$, and $b = 1$. Use the Runge-Kutta method to find a numerical solution. Plot the solution to this differential equation for $p_0 = 80$. What is the long-term behavior of the fish population? 
```{r}
# Type your code here (and delete these comments)
```

## Exercise 4

Consider the initial-value problem $\displaystyle\frac{dy}{dt}=\frac{1}{(y+1)(t-2)}$ with $y(-2)=0$. The analytical solution for this IVP is
\[
y(t) = -1 + \sqrt{1+\ln \left(\frac{(t-2)^2}{16}\right)}
\]

(a) Plot the analytical solution for the IVP over the interval $t=-2$ to $t=2$. Add to the plot numerical solutions using Euler's Method, with the number of steps $n=64,~128,~256$. Adjust the y-axis limits as needed.
```{r}
# Type your code here (and delete these comments)
```
(b) Again, plot the analytical solution over the interval $t=-2$ to $t=2$. Now add, the 4th-order Runga-Kutta solutions using $n=100$ and $200$. Adjust the y-axis limits as needed.
```{r}
# Type your code here (and delete these comments)
```
(c) Comment on your observations of the previous two parts. Do you have any explanation for your observations?